---
title: "Credit Data Processing"
author: "Minh Duc Nguyen"
date: "2025-04-06"
output:
  word_document: default
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(data.table)
library(lubridate)
library(dplyr)  # used only for model.matrix
```

```{r config}
DATA_DIR     <- "E:/NEU SLIDE/DATA SCIENCE"
INPUT_FILE   <- file.path(DATA_DIR, "credit.csv")
INTERIM_FILE <- file.path(DATA_DIR, "new credit.csv")
OUTPUT_FILE  <- file.path(DATA_DIR, "credit_cleaned.csv")
```

# Helper Functions

```{r helpers}
# Encode binary flag: "Y"/"N" -> 1/0, preserve NA
encode_binary <- function(x, yes_val = "Y", no_val = "N") {
  x[x == ""] <- NA
  fifelse(is.na(x), NA_integer_,
          fifelse(x == yes_val, 1L, 0L))
}

# Encode ordered factor -> integer rank, preserve NA
encode_ordered <- function(x, levels_vec) {
  x[x == ""] <- NA
  match(x, levels_vec)
}

# Parse "Mon-YYYY" date strings -> POSIXct, preserve NA
parse_mon_year <- function(x) {
  x[x == ""] <- NA
  parse_date_time(x, orders = "b-Y")
}

# Compute days elapsed since issue_d (vectorized)
days_since_issue <- function(date_col, issue_d) {
  ifelse(is.na(date_col) | is.na(issue_d), NA_real_,
         as.numeric(date_col - issue_d))
}

# Merge individual + joint columns; joint takes priority when application_type == 1
merge_joint <- function(dt, ind_col, joint_col, new_col) {
  dt[, (new_col) := fifelse(
    application_type == 1 & !is.na(get(joint_col)),
    get(joint_col),
    get(ind_col)
  )]
  dt[, c(ind_col, joint_col) := NULL]
  invisible(dt)
}

# Print sorted missing value report
missing_report <- function(dt, label = "") {
  pct <- sort(colMeans(is.na(dt)) * 100, decreasing = TRUE)
  pct <- pct[pct > 0]
  cat("\n--- Missing report", label, "---\n")
  cat("Variables with missing:", length(pct), "\n")
  print(round(pct, 2))
  invisible(pct)
}
```

# Part 1 — Variable Encoding

## 1. Load Data

```{r load-raw}
credit <- fread(INPUT_FILE)
cat("Raw data:", nrow(credit), "rows,", ncol(credit), "columns\n")
```

## 2. Drop Non-informative Identifiers

Columns such as `id`, `url`, `desc` carry no predictive signal and are removed.

```{r drop-ids}
drop_id_cols <- c("id", "member_id", "url", "desc", "title",
                  "emp_title", "purpose", "zip_code", "addr_state")
credit[, (intersect(drop_id_cols, names(credit))) := NULL]
```

## 3. Term

Strip `" months"` suffix and convert to integer (36 or 60).

```{r term}
credit[, term := as.integer(gsub(" months", "", term))]
```

## 4. Grade & Sub-grade

Ordered risk encoding: A = lowest risk (1), G = highest risk (7).
Sub-grades run A1 (1) through G5 (35).

```{r grade}
grade_levels     <- c("A","B","C","D","E","F","G")
sub_grade_levels <- paste0(rep(grade_levels, each = 5), rep(1:5, times = 7))

credit[, grade     := encode_ordered(grade,     grade_levels)]
credit[, sub_grade := encode_ordered(sub_grade, sub_grade_levels)]
```

## 5. Employment Length

Map text categories to integer years (0–10).

```{r emp-length}
emp_map <- c("< 1 year" = 0L, "1 year"  = 1L, "2 years" = 2L,
             "3 years"  = 3L, "4 years" = 4L, "5 years" = 5L,
             "6 years"  = 6L, "7 years" = 7L, "8 years" = 8L,
             "9 years"  = 9L, "10+ years" = 10L)
credit[emp_length == "", emp_length := NA_character_]
credit[, emp_length := emp_map[emp_length]]
```

## 6. Home Ownership

One-hot encode via `model.matrix`. Missing values are treated as a separate `"Missing"` category.

```{r home-ownership}
credit[home_ownership == "", home_ownership := NA_character_]
credit[, home_ownership_temp := fifelse(is.na(home_ownership), "Missing", home_ownership)]
home_dummies <- model.matrix(~ home_ownership_temp - 1,
                             data = as.data.frame(credit[, .(home_ownership_temp)]))
credit <- cbind(credit, as.data.table(home_dummies))
credit[, c("home_ownership", "home_ownership_temp") := NULL]
```

## 7. Issue Date

Parse to `POSIXct` — used later as the reference point for all relative date features.

```{r issue-date}
credit[, issue_d := parse_mon_year(issue_d)]
```

## 8. Loan Status

Ordered risk encoding: Fully Paid = 1 (lowest risk) → Default = 9 (highest risk).

```{r loan-status}
status_levels <- c(
  "Fully Paid", "Current", "In Grace Period",
  "Late (16-30 days)", "Late (31-120 days)",
  "Does not meet the credit policy. Status:Fully Paid",
  "Does not meet the credit policy. Status:Charged Off",
  "Charged Off", "Default"
)
credit[, loan_status := encode_ordered(loan_status, status_levels)]
```

## 9. Binary Flags

```{r binary-flags}
credit[, pymnt_plan           := encode_binary(pymnt_plan,           "y",         "n")]
credit[, initial_list_status  := encode_binary(initial_list_status,  "w",         "f")]
credit[, hardship_flag        := encode_binary(hardship_flag,        "Y",         "N")]
credit[, application_type     := encode_binary(application_type,     "Joint App", "Individual")]
credit[, disbursement_method  := encode_binary(disbursement_method,  "Cash",      "DirectPay")]
credit[, debt_settlement_flag := encode_binary(debt_settlement_flag, "Y",         "N")]
```

## 10. Date Columns

Parse all remaining date columns to `POSIXct`.

```{r parse-dates}
date_cols <- c("earliest_cr_line", "last_pymnt_d", "next_pymnt_d",
               "last_credit_pull_d", "sec_app_earliest_cr_line",
               "payment_plan_start_date", "debt_settlement_flag_date",
               "settlement_date")
for (col in intersect(date_cols, names(credit))) {
  credit[, (col) := parse_mon_year(get(col))]
}
```

## 11. Hardship Type

Derive a binary flag; `1` = interest-only deferral exists, `0` otherwise.

```{r hardship-type}
credit[, has_hardship := fifelse(
  is.na(hardship_type), NA_integer_,
  fifelse(hardship_type == "INTEREST ONLY-3 MONTHS DEFERRAL", 1L, 0L)
)]
credit[, hardship_type := NULL]
```

## 12. Hardship Reason

One-hot encode via `model.matrix`. Missing values treated as `"Missing"` category.

```{r hardship-reason}
credit[hardship_reason == "", hardship_reason := NA_character_]
credit[, hardship_reason_temp := fifelse(is.na(hardship_reason), "Missing", hardship_reason)]
hardship_dummies <- model.matrix(~ hardship_reason_temp - 1,
                                 data = as.data.frame(credit[, .(hardship_reason_temp)]))
credit <- cbind(credit, as.data.table(hardship_dummies))
credit[, c("hardship_reason", "hardship_reason_temp") := NULL]
```

## 13. Ordered Status Variables

```{r ordered-status}
credit[, hardship_status := encode_ordered(
  hardship_status, c("COMPLETED", "ACTIVE", "BROKEN"))]
credit[, hardship_loan_status := encode_ordered(
  hardship_loan_status, c("Current", "In Grace Period",
                          "Late (16-30 days)", "Late (31-120 days)", "Issued"))]
credit[, settlement_status := encode_ordered(
  settlement_status, c("COMPLETED", "ACTIVE", "BROKEN"))]
```

## 14. Hardship Dates → `ever_hardship` Flag

```{r ever-hardship}
credit[, ever_hardship := fifelse(
  hardship_start_date == "" | hardship_start_date == "0000-00-00", 0L, 1L
)]
credit[, c("hardship_start_date", "hardship_end_date") := NULL]
```

## 15. Payment Plan → `has_payment_plan` Flag

```{r payment-plan}
credit[, has_payment_plan := fifelse(
  is.na(payment_plan_start_date), 0L, 1L
)]
```

## 16. Verification Status

Ordered encoding: Not Verified = 1, Source Verified = 2, Verified = 3.

```{r verification}
verif_levels <- c("Not Verified", "Source Verified", "Verified")
credit[, verification_status       := encode_ordered(verification_status,       verif_levels)]
credit[, verification_status_joint := encode_ordered(verification_status_joint, verif_levels)]
```

## 17. Date Columns → Days Since Issue

Convert all date columns to days elapsed since `issue_d`, then drop originals.

```{r days-since-issue}
date_cols_relative <- c("earliest_cr_line", "last_pymnt_d", "next_pymnt_d",
                        "last_credit_pull_d", "sec_app_earliest_cr_line",
                        "payment_plan_start_date", "debt_settlement_flag_date",
                        "settlement_date")
date_cols_relative <- intersect(date_cols_relative, names(credit))
for (col in date_cols_relative) {
  credit[, (paste0(col, "_since_issue")) := days_since_issue(get(col), issue_d)]
}
credit[, c(date_cols_relative, "issue_d") := NULL]
```

## 18. Final Check

```{r check-encoding}
non_numeric <- names(credit)[!sapply(credit, is.numeric)]
cat("Non-numeric columns remaining:", length(non_numeric), "\n")
if (length(non_numeric) > 0) print(non_numeric)
```

## 19. Save Interim File

```{r save-interim}
fwrite(credit, INTERIM_FILE)
cat("Saved:", INTERIM_FILE, "\n")
cat("Size:", nrow(credit), "rows,", ncol(credit), "columns\n")
```

# Part 2 — Missing Value Handling

## 1. Load Interim File

```{r load-interim}
credit <- fread(INTERIM_FILE)
cat("Loaded:", nrow(credit), "rows,", ncol(credit), "columns\n")
if ("V1" %in% names(credit)) credit[, V1 := NULL]
```

## 2. Merge Individual / Joint Columns

When `application_type == 1` (Joint App), the joint value takes priority.

| New column | Individual source | Joint source |
|---|---|---|
| `income_used` | `annual_inc` | `annual_inc_joint` |
| `dti_used` | `dti` | `dti_joint` |
| `revol_bal_used` | `revol_bal` | `revol_bal_joint` |
| `verification_status_used` | `verification_status` | `verification_status_joint` |

```{r merge-joint}
merge_joint(credit, "annual_inc",          "annual_inc_joint",          "income_used")
merge_joint(credit, "dti",                 "dti_joint",                 "dti_used")
merge_joint(credit, "revol_bal",           "revol_bal_joint",           "revol_bal_used")
merge_joint(credit, "verification_status", "verification_status_joint", "verification_status_used")
```

## 3. Missing Value Report

```{r missing-report}
missing_report(credit, "after joint merge")
```

## 4. Three-Tier Missing Strategy

| Missing rate | Action | Rationale |
|---|---|---|
| ≤ 5% | Drop rows | Safe to remove — minimal data loss |
| 5% – 50% | Drop columns | Too sparse to impute reliably |
| > 50% | Drop columns | Majority missing — no useful signal |

### 4a. Drop Rows (≤ 5% missing)

```{r drop-rows}
drop_rows_vars <- intersect(c(
  "loan_amnt", "income_used", "dti_used", "delinq_2yrs",
  "inq_last_6mths", "revol_util", "collections_12_mths_ex_med",
  "tot_coll_amt", "bc_open_to_buy", "bc_util",
  "mo_sin_old_il_acct", "mths_since_recent_bc",
  "num_rev_accts", "num_tl_120dpd_2m", "pct_tl_nvr_dlq",
  "last_pymnt_d_since_issue", "last_credit_pull_d_since_issue"
), names(credit))

before <- nrow(credit)
credit <- na.omit(credit, cols = drop_rows_vars)
cat("Dropped rows:", before - nrow(credit), "| Remaining:", nrow(credit), "\n")
```

### 4b. Drop Columns (5–50% missing)

```{r drop-cols-medium}
drop_cols_medium <- intersect(c(
  "open_acc_6m", "open_act_il", "open_il_12m", "open_il_24m",
  "mths_since_rcnt_il", "total_bal_il", "il_util",
  "open_rv_12m", "open_rv_24m", "max_bal_bc", "all_util",
  "inq_fi", "total_cu_tl", "inq_last_12m",
  "mths_since_recent_inq", "emp_length"
), names(credit))
credit[, (drop_cols_medium) := NULL]
cat("Dropped columns (5-50%):", length(drop_cols_medium), "\n")
```

### 4c. Drop Columns (> 50% missing)

```{r drop-cols-high}
drop_cols_high <- intersect(c(
  "mths_since_last_delinq", "mths_since_recent_revol_delinq",
  "mths_since_recent_bc_dlq", "mths_since_last_record",
  "mths_since_last_major_derog",
  "sec_app_fico_range_low", "sec_app_fico_range_high",
  "sec_app_inq_last_6mths", "sec_app_mort_acc", "sec_app_open_acc",
  "sec_app_revol_util", "sec_app_open_act_il", "sec_app_num_rev_accts",
  "sec_app_chargeoff_within_12_mths",
  "sec_app_collections_12_mths_ex_med",
  "sec_app_mths_since_last_major_derog",
  "deferral_term", "hardship_amount", "hardship_length", "hardship_dpd",
  "orig_projected_additional_accrued_interest",
  "hardship_payoff_balance_amount", "hardship_last_payment_amount",
  "settlement_amount", "settlement_percentage", "settlement_term",
  "hardship_status", "hardship_loan_status", "settlement_status",
  "next_pymnt_d_since_issue", "sec_app_earliest_cr_line_since_issue",
  "payment_plan_start_date_since_issue",
  "debt_settlement_flag_date_since_issue",
  "settlement_date_since_issue", "lnext_pymnt_d_since_issue"
), names(credit))
credit[, (drop_cols_high) := NULL]
cat("Dropped columns (>50%):", length(drop_cols_high), "\n")
```

## 5. Final Check

```{r final-check}
cat("Total missing values remaining:", sum(is.na(credit)), "\n")
cat("Final size:", nrow(credit), "rows,", ncol(credit), "columns\n")
```

## 6. Save Output

```{r save-output}
fwrite(credit, OUTPUT_FILE)
cat("Saved:", OUTPUT_FILE, "\n")
```

## 7. Free RAM

```{r free-ram}
rm(list = ls())
gc()
cat("RAM cleared\n")
```
